---
title: "Get started with datey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with datey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datey)
```

This vignette is a hands-on guide to the **datey** package. For the
motivation behind the annual-grid approach and the associativity guarantee,
see [Why **datey**?][why-datey]. For the complete formal specification, see the
[**datey** specification][spec].

## Core types

**datey** provides three S3 classes:

- **`datey`** -- a point in time, stored at day-fraction precision.
- **`durationy`** -- a duration in years.
- **`datey_interval`** -- a half-open `[start, end)` time interval.

These are atomic types[^datey-interval-is-atomic] that store dates and durations as 
integers with units of 1/534&#8201;360 of a
year). As a result, arithmetic with these types is exact and associative.

[^datey-interval-is-atomic]: Even though `datey_interval` stores the start and end of
a time interval, it too is atomic, which means that `datey_interval`s can be
stored in a single R vector without any additional special handling.

## Creating a `datey`

### When in the day?

Exposure periods specified by the dates \eqn{a} to \eqn{b} typically mean that
the whole of the day \eqn{a} and the whole of the day \eqn{b} are included.
In the **datey** system this corresponds to using 
`start_day()` for \eqn{a} and
`end_day()` for \eqn{b}.

Deaths on the other hand typically happen *during* a day.
In the **datey** system this corresponds to using 
`mid_day()`.

These distinctions may be new to you and your first reaction may be that they
are immaterial. But it costs very little to be precise and sometimes
systematic errors can accumulate and end up being material.

### From year, month and day

`start_day()`, `mid_day()` and `end_day()` create a `datey` from scratch:

```{r}
start_day(2024, 3, 7)   # Start of the day 7 March 2024
mid_day(2024, 3, 7)     # Middle of the day 7 March 2024
end_day(2024, 3, 7)     # End of the day 7 March 2024
```

The end of a day is the same point as the start of the next, so
`end_day()` applied to a day is identical to `start_day()` applied to the
following day:

```{r}
identical(end_day(2024, 3, 7), start_day(2024, 3, 8))
```

For an arbitrary position within a day, `datey()` accepts a day fraction
between 0 and 1:

```{r}
datey(2024, 3, 7, 0.25)   # Quarter of the way through the day 7 March 2024
```

### From base R dates

It is often the case that data already contains dates defined using the
standard base R types `Date`[^fractional-dates], `POSIXct` or `POSIXlt`.

To convert these to a `datey`, use `start_day()`, `mid_day()` or `end_day()`:

```{r}
d <- as.Date("2024-03-07")
start_day(d)
mid_day(d)
end_day(d)
```

[^fractional-dates]: Even though the `Date` type is not designed for
fractional dates, it typically uses floating point under the covers,
and can unintentionally end up with a fractional value
e.g. by taking a mean of `Date`s. For this reason, 
a `day_fraction` argument is always required for a `Date`.

### From fractional years or text

`datey()` also accepts a fractional calendar year or a character string in
`YYYY-MM-DD[.f]` format:

```{r}
datey(2024)           # Start of calendar year 2024
datey(2024.5)         # Midway through calendar year 2024

datey("2024-03-07")   # Start of the day 7 March 2024 (day fraction defaults to 0)
datey("2024-03-07.5") # Middle of the day 7 March 2024
```

## Properties of a `datey`

The `$` operator extracts components of a `datey`:

```{r}
t <- mid_day(2024, 3, 7)
t$year
t$month
t$day
t$day_fraction
```

If you need several components at once, it is more efficient to use 
`to_ymdf()` instead:

```{r}
to_ymdf(t)
```

`as.double()` converts to a fractional calendar year; `as.integer()` gives
the calendar year:

```{r}
as.double(t)
as.integer(t)
```

`is_start_day()` and `is_mid_day()` test the position within the day. Note
that `end_day()` produces a `datey` at the start of the following day, so it
tests as `is_start_day()`:

```{r}
is_start_day(start_day(2024, 3, 7)) # TRUE
is_mid_day(mid_day(2024, 3, 7))     # TRUE
is_start_day(end_day(2024, 3, 7))   # TRUE because end = start of next day
is_mid_day(datey(2024, 3, 7, 0.25)) # FALSE
```

## Creating a `durationy`

`durationy`s typically arise as `datey` differences:

```{r}
dob <- start_day(as.Date("1965-09-12"))
dod <- mid_day(2024, 3, 7)
age <- dod - dob
age
```

You can create them explicitly using `durationy()`, which accepts a number of
years:

```{r}
durationy(1)      # One year
durationy(0.5)    # Half a year
durationy(-2.5)   # Two and a half years in the past
```

And you can convert them back to numerics using
`as.double()`, which gives the duration as years, and
`as.integer()`, which truncates toward
zero:

```{r}
as.double(age)
as.integer(age)   # Whole years only
```

## Comparisons and arithmetic

A number of arithmetic operations are available for `datey`, `durationy` and
`datey_interval`.

Beware that not all combinations are valid because, for instance,
it doesn't make sense to add two dates together.

The table below summarises the valid arithmetic and comparison operations.
All arithmetic is carried out as exact integer arithmetic on the underlying
click counts, so the results are exact and associative.

| Left | Op | Right | Result |
|:---|:---|:---|:---|
| `datey` | `-` | `datey` | `durationy` |
| `datey` | `+ -` | `durationy` | `datey` |
| `durationy` | `+` | `datey` | `datey` |
| `durationy` | `+ -` | `durationy` | `durationy` |
| `datey` | `== != < <= > >=` | `datey` | `logical` |
| `durationy` | `== != < <= > >=` | `durationy` | `logical` |
| `datey` | `%to%` | `datey` | `datey_interval` |
| `datey_interval` | `== !=` | `datey_interval` | `logical` |
| `datey_interval` | `%includes%` | `datey` | `logical` |
| `datey_interval` | `&` | `datey_interval` | `datey_interval` |

```{r}
start  <- start_day(2000, 1, 1)
one_yr <- durationy(1)
qtr_yr <- durationy(0.25)

start + one_yr    # One year later
start - qtr_yr    # Quarter of a year earlier

one_yr - qtr_yr   # Three quarters of a year
one_yr + qtr_yr

datey(2024) < datey(2025)     # TRUE
durationy(1) > durationy(0.5) # TRUE
```

You can also do mixed arithmetic with `datey` and `durationy` and numbers,
in which case `datey`s and `durationy`s are first converted to `double`s:

```{r}
identical(datey(2000) + 25, 2025)     # TRUE
identical(durationy(2) * 0.05, 0.10)  # TRUE
```

## `datey_interval` -- representing a time period

A `datey_interval` is a half-open `[start, end)` interval. Create one with
`datey_interval()` or the `%to%` operator:

```{r}
a  <- start_day(2024, 1, 1)
b  <- start_day(2025, 1, 1)
interval <- a %to% b
interval
```

The `$start`, `$end` and `$duration` properties extract the interval's
components:

```{r}
interval$start
interval$end
interval$duration
```

`durationy()` accepts a `datey_interval` directly:

```{r}
durationy(interval)
```

### Interval membership testing

`%includes%` tests whether a `datey` falls inside the interval. The interval
includes its start and excludes its end:

```{r}
interval %includes% a                     # TRUE  -- start is included
interval %includes% b                     # FALSE -- end is excluded
interval %includes% mid_day(2024, 6, 15)  # TRUE
```

### Interval properties

`is_proper()` returns `TRUE` when start ≤ end; `is_collapsed()` returns
`TRUE` when start ≥ end. A point interval `[a, a)` is both proper and
collapsed (it contains no time):

```{r}
is_proper(interval)    # TRUE because start <= end
is_collapsed(interval) # FALSE because start < end

point <- a %to% a      # Empty (point) interval
is_proper(point)       # TRUE because a <= a
is_collapsed(point)    # TRUE because a >= a
```

### Intersection

The `&` operator returns the intersection of two `datey_interval`s. This is
the most direct way to compute the overlap of two time periods:

```{r}
period    <- start_day(2023, 7, 1) %to% end_day(2024, 6, 30)
year_2024 <- start_day(2024, 1, 1) %to% end_day(2024, 12, 31)

overlap <- period & year_2024
overlap

overlap$duration  # exposure in calendar year 2024, in years
```

## NA values

Throughout the **datey** package, `NA` will cause an error when used where
a `datey_`, `durationy_` or `datey_interval_` is expected.
This is because the type of `NA` is `logical`. which has no meaningful 
date or duration interpretation therefore potentially indicates user
error.

If you want an NA value with a **datey** system type, use the explicit forms
`NA_datey_`, `NA_durationy_` or `NA_datey_interval_` as appropriate.

`is.na()` and `anyNA()` work as expected:

```{r}
is.na(NA_datey_)
anyNA(c(datey(2000), NA_datey_, datey(2024)))
```

By default, out-of-range inputs stop execution. With
`strict = FALSE` they become `NA` instead:

```{r}
datey(999.9, strict = FALSE)           # Outside [1000,3000]: NA
start_day(2000, 0, 12, strict = FALSE) # Invalid month: NA
mid_day(2001, 2, 29, strict = FALSE)   # Invalid day (given year and month): NA
durationy(2000.1, strict = FALSE)      # exceeds 2000-year limit: NA
```

NA values propagate through arithmetic:

```{r}
start_day(2024, 1, 1) + NA_durationy_
```

## Sequences and statistics

`seq()`, `min()`, `max()`, `range()` and `mean()` all work on `datey` and
`durationy` vectors:

```{r}
dates <- c(datey(2021), datey(2022), datey(2023))

min(dates)
max(dates)
mean(dates)

seq(from = datey(2020), to = datey(2024), by = durationy(2))
```

[why-datey]: why-datey.html
[spec]: spec.html
