---
title: "Get started with datey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with datey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datey)
```

This vignette is a hands-on guide to the **datey** package. For the
motivation behind the annual-grid approach and the associativity guarantee,
see [Why **datey**?][why-datey]. For the complete formal specification, see the
[**datey** specification][spec].

## The three types

**datey** provides three S3 classes:

- **`datey`** -- a point in time, stored at day-fraction precision.
- **`durationy`** -- a duration in years.
- **`datey_interval`** -- a half-open `[start, end)` time interval.

All values are stored as exact integers (in units of 1/534&#8201;360 of a
year), so arithmetic is both exact and associative.

## Creating a `datey`

### From year, month and day

`start_day()`, `mid_day()` and `end_day()` position a `datey` at the start,
middle or end of a calendar day:

```{r}
start_day(2024, 3, 7)   # start of 7 March 2024
mid_day(2024, 3, 7)     # middle of 7 March 2024
end_day(2024, 3, 7)     # end of 7 March 2024
```

The end of a day is the same point as the start of the next, so
`end_day()` applied to a day is identical to `start_day()` applied to the
following day:

```{r}
identical(end_day(2024, 3, 7), start_day(2024, 3, 8))
```

For an arbitrary position within a day, `from_ymdf()` accepts a day fraction
between 0 and 1:

```{r}
from_ymdf(2024, 3, 7, 0.25)   # one quarter of the way through 7 March 2024
```

### From base R dates

`as_start_day()`, `as_mid_day()` and `as_end_day()` convert `Date`,
`POSIXct` or `POSIXlt` values, fixing the day fraction:

```{r}
d <- as.Date("2024-03-07")
as_start_day(d)
as_mid_day(d)
as_end_day(d)
```

A `Date` carries no time component, so `datey(d)` is equivalent to
`as_start_day(d)`.

### From fractional years or text

`datey()` also accepts a fractional calendar year or a character string in
`YYYY-MM-DD[.f]` format:

```{r}
datey(2024)     # start of 2024
datey(2024.5)   # halfway through 2024

datey("2024-03-07")   # start of 7 March 2024 (day fraction defaults to 0)
datey("2024-03-07.5") # middle of 7 March 2024
```

## Properties of a `datey`

The `$` operator extracts components of a `datey`:

```{r}
t <- mid_day(2024, 3, 7)
t$year
t$month
t$day
t$day_fraction
```

When several components are needed at once, `to_ymdf()` is more efficient:

```{r}
to_ymdf(t)
```

`as.double()` converts to a fractional calendar year; `as.integer()` gives
the calendar year:

```{r}
as.double(t)
as.integer(t)
```

`is_start_day()` and `is_mid_day()` test the position within the day. Note
that `end_day()` produces a `datey` at the start of the following day, so it
tests as `is_start_day()`:

```{r}
is_start_day(start_day(2024, 3, 7))     # TRUE
is_mid_day(mid_day(2024, 3, 7))         # TRUE
is_start_day(end_day(2024, 3, 7))       # TRUE because end = start of next day
is_mid_day(from_ymdf(2024, 3, 7, 0.25)) # FALSE
```

## Creating a `durationy`

`durationy()` accepts a number of years:

```{r}
durationy(1)      # one year
durationy(0.5)    # half a year
durationy(-2.5)   # two and a half years in the past
```

The most natural way to obtain a `durationy` is to subtract two `datey`
values:

```{r}
dob <- as_start_day(as.Date("1965-09-12"))
dod <- mid_day(2024, 3, 7)
age <- dod - dob
age
```

`as.double()` gives the duration as years; `as.integer()` truncates toward
zero:

```{r}
as.double(age)
as.integer(age)   # whole years only
```

## Arithmetic

The table below summarises the typed arithmetic operations. All arithmetic is
carried out as exact integer arithmetic on the underlying click counts, so the
results are exact and associative.

| Left | Op | Right | Result |
|:---|:---|:---|:---|
| `datey` | `-` | `datey` | `durationy` |
| `datey` | `+ -` | `durationy` | `datey` |
| `durationy` | `+` | `datey` | `datey` |
| `durationy` | `+ -` | `durationy` | `durationy` |
| `datey` | `== != < <= > >=` | `datey` | logical |
| `durationy` | `== != < <= > >=` | `durationy` | logical |

```{r}
start  <- start_day(2000, 1, 1)
one_yr <- durationy(1)
qtr_yr <- durationy(0.25)

start + one_yr    # one year later
start - qtr_yr    # a quarter of a year earlier

one_yr - qtr_yr   # three quarters of a year
one_yr + qtr_yr

datey(2024) < datey(2025)       # TRUE
durationy(1) > durationy(0.5)   # TRUE
```

## `datey_interval` -- representing a time period

A `datey_interval` is a half-open `[start, end)` interval. Create one with
`datey_interval()` or the `%to%` operator:

```{r}
a  <- start_day(2024, 1, 1)
b  <- start_day(2025, 1, 1)
interval <- a %to% b
interval
```

The `$start`, `$end` and `$duration` properties extract the interval's
components:

```{r}
interval$start
interval$end
interval$duration
```

`durationy()` accepts a `datey_interval` directly:

```{r}
durationy(interval)
```

### Membership testing

`%includes%` tests whether a `datey` falls inside the interval. The interval
includes its start and excludes its end:

```{r}
interval %includes% a                     # TRUE  -- start is included
interval %includes% b                     # FALSE -- end is excluded
interval %includes% mid_day(2024, 6, 15)  # TRUE
```

### Interval properties

`is_proper()` returns `TRUE` when start ≤ end; `is_collapsed()` returns
`TRUE` when start ≥ end. A point interval `[a, a)` is both proper and
collapsed (it contains no time):

```{r}
is_proper(interval)    # TRUE because start <= end
is_collapsed(interval) # FALSE because start < end

pt <- a %to% a         # empty (point) interval
is_proper(pt)          # TRUE because a <= a
is_collapsed(pt)       # TRUE because a >= a
```

### Intersection

The `&` operator returns the intersection of two `datey_interval`s. This is
the most direct way to compute the overlap of two time periods:

```{r}
period    <- start_day(2023, 7, 1) %to% end_day(2024, 6, 30)
year_2024 <- start_day(2024, 1, 1) %to% end_day(2024, 12, 31)

overlap <- period & year_2024
overlap

overlap$duration  # exposure in calendar year 2024, in years
```

## NA values

Each type has a dedicated NA constant:

```{r}
NA_datey_
NA_durationy_
NA_datey_interval_
```

`is.na()` and `anyNA()` work as expected:

```{r}
is.na(NA_datey_)
anyNA(c(datey(2000), NA_datey_, datey(2024)))
```

By default (`strict = TRUE`), out-of-range inputs stop execution. With
`strict = FALSE` they become `NA` instead:

```{r}
datey(999.9, strict = FALSE)     # below valid range: NA
durationy(2001, strict = FALSE)  # exceeds 2000-year limit: NA
```

NA values propagate through arithmetic:

```{r}
start_day(2024, 1, 1) + NA_durationy_
```

## Sequences and statistics

`seq()`, `min()`, `max()`, `range()` and `mean()` all work on `datey` and
`durationy` vectors:

```{r}
dates <- c(datey(2021), datey(2022), datey(2023))

min(dates)
max(dates)
mean(dates)

seq(from = datey(2020), to = datey(2024), by = durationy(2))
```

[why-datey]: why-datey.html
[spec]: spec.html
