---
title: "Get started with datey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with datey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datey)
```

Link to other pages here


When converting dates, which are measured in days, to a system that assumes all years have the same length, something has to give. 

The most common approach, which simply because some years contain 365 days and others have 366. 


 (which is to assume that days are 1 year / N, where N is the number of days in the containing calendar year)
- a consistent framework based on the most common approach,


- Fixed precision dates for calendar years [1000, 3000)
- Technically this is the Proleptic Gregorian calendar

Guarantees:

- Can represent start and middle of any day leap year and non leap year
- Years divisible by 120

You can convert from standard base R dates (`Date`, `POSIXct` and `POSIXlt`).

Key argument is `day_fraction`. Simpler just to use `start_day`, `mid_day` or 
`end_day`.

For instance

- start of E2R is typically included, so use `start_day`.
- if the individual survived then the end of the E2R is also typically included, 
so use `end_day`. This means that if the E2R comprises separate periods then 
they will interlock correctly (E2Rs being [measurable](https://timgord.com/2025-08/mortality-measures-matter/#Insight2))
- If the individual died then on average this is during the final day and so 
use `mid_day`.


Guarantees:

- Can represent start and middle of any day leap year and non leap year
- Years divisible by 120

You can convert from standard base R dates (`Date`, `POSIXct` and `POSIXlt`).

Key argument is `day_fraction`. Simpler just to use `start_day`, `mid_day` or 
`end_day`.

For instance

- start of E2R is typically included, so use `start_day`.
- if the individual survived then the end of the E2R is also typically included, 
so use `end_day`. This means that if the E2R comprises separate periods then 
they will interlock correctly (E2Rs being [measurable](https://timgord.com/2025-08/mortality-measures-matter/#Insight2))
- If the individual died then on average this is during the final day and so 
use `mid_day`.
